---
title: "LB_7_ASV_heatmap"
author: "Laura Brettell"
date: "2022-11-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The project

Vertical transmission – isofemale lines and rearing in closed conditions free from external microbes to examine transfer of bacteria between generations 

## About

This script uses the rarefied ASV table made previously, filters to keep only most abundant ASVs and generates heatmaps to show relative abundance of top ASVs across sampletypes


using info from https://www.r-bloggers.com/2013/02/from-otu-table-to-heatmap/



# Install packages

```{r}

# phyloseq itself
if(!requireNamespace("BiocManager")){
  install.packages("BiocManager")
   }
BiocManager::install("phyloseq")

# qiime2r
# to be able to convert my qiime2 files
if (!requireNamespace("devtools", quietly = TRUE)){install.packages("devtools")}
 devtools::install_github("jbisanz/qiime2R") # current version is 0.99.20
 
# decontam for filtering contaminant reads
# if (!require("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")

#BiocManager::install("decontam")

#install.packages("MetBrewer")


#library(MetBrewer)
library(phyloseq)
library(qiime2R)
#library(ggplot2)
library(tidyverse) # for filter_all()
#library(dplyr) # for select command

#library(gridExtra) # to arrange heatmaps


```
# import data


```{r}
# get metadata
metadata <- read_q2metadata("metadata_updated_01092022_NAs_removed_new_cols.txt") # this is everything including those samples which were removed after rarefaction etc

# get taxa
ASVs <- read_qza("rarefied_table.qza") # now read in ASV data
ASVs <- as.data.frame(ASVs$data)

# get taxonomy info
taxonomydata <- read_qza("taxonomy-silva.qza")$data # read in data
tax_parse <- parse_taxonomy(taxonomydata)# make in more useful format

```

```{r}
# make ASV table into proportions
ASV_proportion <- ASVs
ASV_proportion <- scale(ASV_proportion, center=F, scale=colSums(ASV_proportion))

```

# filtering ASVs to only the more common in the dataset

keeping all ASVs present at >= 5% reads in >=1 sample

```{r}

ASV_subset <- filter_all(as.data.frame(ASV_proportion), any_vars(. >= 0.05))
# reduces from 4516 ASVs to 125
ASV_subset_transpose_rownames <- -t(ASV_subset)
ASV_subset_transpose_rownames <- as.data.frame(ASV_subset_transpose_rownames)
ASV_subset_transpose <- rownames_to_column(ASV_subset_transpose_rownames, var = "SampleID")

# now to subset the taxonomy data to keep only the taxonomy for the ASV subset
ASV_subset_names <- row.names(ASV_subset)
tax_subset <- tax_parse[ASV_subset_names,]

```


```{r}
# now to get an extra taxonomy column with the full long name and make unique for each ASV - to help with plotting
# get the unformatted taxonomy info. This is useful for adding info to the plot later
taxonomy_long <- read_qza("taxonomy-silva.qza")$data
taxonomy_long <- taxonomy_long[,c(1:2)]
# just need featureid and taxon

temp <- tax_subset
#temp <- as.data.frame(otu_table(ps.adult.offspring.filt))
temp$Feature.ID <- row.names(temp) # specifying a col with FeatureID to merge with tax table

temp <- merge(temp, taxonomy_long, by = "Feature.ID")
# i want to call them by their tazon, but there are multiple ASVs witht he same taxon so to give each row a unique name I must ammend to have the featureID and taxon together
temp$longname <- paste(temp$Feature.ID, "-", temp$Taxon)
temp <- column_to_rownames(temp, var = "Feature.ID")

#remove the non-numeric columns now
#temp <- subset(temp, select = -c(Feature.ID, Taxon))
#temp <- -t(temp)


```



# put into phyloseq object to make heatmap

(to get phyloseq obj metadata etc)

```{r}
# get taxonomy into the right format
taxmat_subs <- as.matrix(temp) # must be matrix for later steps
TAX = tax_table(taxmat_subs) # convert to necessary format (formal class taxonomyTable) to make phyloseq object

# get ASV table into the right format
ASV_subset2 <- ASV_subset
ASV_subset2[ASV_subset2 == 0] <- NA
ASVmat_subs <- as.matrix(ASV_subset2) # convert to matrix
OTU = otu_table(ASVmat_subs, taxa_are_rows = TRUE) # convert to necessary format  to make phyloseq object

# create phyloseq object
physeq = phyloseq(OTU, TAX) #create phyloseq object

# get metadata matching up correctly
physeq_name_order <- sample_names(physeq) # need to get the samples in the same order as they appear in the physeq object (ie the column orders of the ASVs)

metadata_subset <- merge(metadata, ASV_subset_transpose, by="SampleID") # this way I dont need to filter after as it only keeps rows in both original dfs
metadata_physeq_order <- metadata_subset 
metadata_physeq_order %>%  arrange(SampleID, physeq_name_order)

# make object of a class that phyloseq understands, by cropping the columns of interest from the metadata and specifying the same rows as the original phloseq object
# this has an additional 2 col names compared to original 'sampletype-order' and 'line-order' just made added for easy plotting.

sampledata2 = sample_data(data.frame(
  Sample = metadata_physeq_order$SampleID,
  seqID = metadata_physeq_order$seqid,
  location = metadata_physeq_order$location,
  line = metadata_physeq_order$line,
  sampletype = metadata_physeq_order$sampletype,
  neg_control = metadata_physeq_order$neg_control,
  sample_order = metadata_physeq_order$`sampletype-order`,
  line_order = metadata_physeq_order$`line-order`,
  row.names=sample_names(physeq),
  stringsAsFactors=FALSE
))

#make phyloseq obj for plotting
physeq2 = merge_phyloseq(physeq, sampledata2)

saveRDS(physeq2, "physeq_top_ASVs")
# save phyloseq object for using for supp cooccurence heatmaps
```

# make heatmap

```{r}
heatmap_test <- plot_heatmap(
              physeq2, 
              method = NULL,
              taxa.label = "longname",
             taxa.order = "Genus",
             sample.order = "sample_order",
             sample.label = "sampletype",
             trans = NULL,
             low = "#B6BCE3",
             high = "#3A3F64",
             na.value = "white",
             title = "heatmap")
             
heatmap_test2 <- heatmap_test + facet_wrap(~line_order, scales = "free_x", nrow = 1)

heatmap_test2

#ggsave("./plots/facet_heatmap_order.pdf", height=8, width=11, device="pdf") 

# an alternative low = "#e7bcbc",high = "#000033",

```

# below is a tester to see how the heatmap looks when reduced to a smaller number of ASVs



# filtering ASVs to only the more common in the dataset

keeping all ASVs present at >= 5% reads in >=1 sample

```{r}

ASV_subset_6perc <- filter_all(as.data.frame(ASV_proportion), any_vars(. >= 0.06))
# reduces from 4516 ASVs to 125
ASV_subset_transpose_rownames <- -t(ASV_subset_6perc)
ASV_subset_transpose_rownames <- as.data.frame(ASV_subset_transpose_rownames)
ASV_subset_transpose <- rownames_to_column(ASV_subset_transpose_rownames, var = "SampleID")

# now to subset the taxonomy data to keep only the taxonomy for the ASV subset
ASV_subset_names <- row.names(ASV_subset)
tax_subset <- tax_parse[ASV_subset_names,]

```


```{r}
# now to get an extra taxonomy column with the full long name and make unique for each ASV - to help with plotting
# get the unformatted taxonomy info. This is useful for adding info to the plot later
taxonomy_long <- read_qza("taxonomy-silva.qza")$data
taxonomy_long <- taxonomy_long[,c(1:2)]
# just need featureid and taxon

temp <- tax_subset
#temp <- as.data.frame(otu_table(ps.adult.offspring.filt))
temp$Feature.ID <- row.names(temp) # specifying a col with FeatureID to merge with tax table

temp <- merge(temp, taxonomy_long, by = "Feature.ID")
# i want to call them by their tazon, but there are multiple ASVs witht he same taxon so to give each row a unique name I must ammend to have the featureID and taxon together
temp$longname <- paste(temp$Feature.ID, "-", temp$Taxon)
temp <- column_to_rownames(temp, var = "Feature.ID")

#remove the non-numeric columns now
#temp <- subset(temp, select = -c(Feature.ID, Taxon))
#temp <- -t(temp)


```



# put into phyloseq object to make heatmap

(to get phyloseq obj metadata etc)

```{r}
# get taxonomy into the right format
taxmat_subs <- as.matrix(temp) # must be matrix for later steps
TAX = tax_table(taxmat_subs) # convert to necessary format (formal class taxonomyTable) to make phyloseq object

# get ASV table into the right format
ASV_subset2 <- ASV_subset
ASV_subset2[ASV_subset2 == 0] <- NA
ASVmat_subs <- as.matrix(ASV_subset2) # convert to matrix
OTU = otu_table(ASVmat_subs, taxa_are_rows = TRUE) # convert to necessary format  to make phyloseq object

# create phyloseq object
physeq = phyloseq(OTU, TAX) #create phyloseq object

# get metadata matching up correctly
physeq_name_order <- sample_names(physeq) # need to get the samples in the same order as they appear in the physeq object (ie the column orders of the ASVs)

metadata_subset <- merge(metadata, ASV_subset_transpose, by="SampleID") # this way I dont need to filter after as it only keeps rows in both original dfs
metadata_physeq_order <- metadata_subset 
metadata_physeq_order %>%  arrange(SampleID, physeq_name_order)

# make object of a class that phyloseq understands, by cropping the columns of interest from the metadata and specifying the same rows as the original phloseq object
# this has an additional 2 col names compared to original 'sampletype-order' and 'line-order' just made added for easy plotting.

sampledata2 = sample_data(data.frame(
  Sample = metadata_physeq_order$SampleID,
  seqID = metadata_physeq_order$seqid,
  location = metadata_physeq_order$location,
  line = metadata_physeq_order$line,
  sampletype = metadata_physeq_order$sampletype,
  neg_control = metadata_physeq_order$neg_control,
  sample_order = metadata_physeq_order$`sampletype-order`,
  line_order = metadata_physeq_order$`line-order`,
  row.names=sample_names(physeq),
  stringsAsFactors=FALSE
))

#make phyloseq obj for plotting
physeq2 = merge_phyloseq(physeq, sampledata2)

#saveRDS(physeq2, "physeq_top_ASVs")
# save phyloseq object for using for supp cooccurence heatmaps
```


# make heatmap

```{r}
heatmap_test <- plot_heatmap(
              physeq2, 
              method = NULL,
              taxa.label = "longname",
             taxa.order = "Genus",
             sample.order = "sample_order",
             sample.label = "sampletype",
             trans = NULL,
             low = "#B6BCE3",
             high = "#3A3F64",
             na.value = "white",
             title = "heatmap")
             
heatmap_test3 <- heatmap_test + facet_wrap(~line_order, scales = "free_x", nrow = 1)

heatmap_test3

#ggsave("./plots/facet_heatmap_order.pdf", height=8, width=11, device="pdf") 

# an alternative low = "#e7bcbc",high = "#000033",

```

