---
title: "VT-filtering-qiime-data"
author: "Laura Brettell"
date: "2022-11-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The project

Vertical transmission – isofemale lines and rearing in closed conditions free from external microbes to examine transfer of bacteria between generations via the egg surface.

## About

This script takes diversity metrics from qiime2 (LB_3_rarefaction_diveristy_qiime2.txt) and makes plots and calculates statistics



# Install packages

```{r}


# to be able to convert my qiime2 files
# if (!requireNamespace("devtools", quietly = TRUE)){install.packages("devtools")}
# devtools::install_github("jbisanz/qiime2R") # current version is 0.99.20
# install.packages("dunn.test")
#install.packages("ggpubr")
#library(devtools)
#install_github("pmartinezarbizu/pairwiseAdonis/pairwiseAdonis")


library(qiime2R)
library(ggplot2)
library(tidyverse)
library(dplyr) # for select command
library(ggpubr)
library(dunn.test)
library(vegan)
library(pairwiseAdonis)


```


# metadata


```{r}
metadata <- read_q2metadata("metadata_updated_01092022_NAs_removed.txt") # this is everything including those samples which were removed after rarefaction etc

```

# alpha diversity - Shannon

Use the shannon diversity data generated in qiime2. 

Get data

```{r}
shannon <- read_qza("shannon_vector.qza")
shannon_meta<-shannon$data %>% rownames_to_column("SampleID") # this moves the sample names to a new column that matches the metadata and allows them to be merged
shannon_meta<-
metadata %>% 
  left_join(shannon_meta)
head(shannon_meta)
# we have now added a column for the shannon diversity to our table which will make everything easy to plot.

shannon_meta <- subset(shannon_meta, shannon_meta$shannon_entropy != "NA")
```

```{r}

shannon_meta %>%
  ggplot+
  geom_boxplot(outlier.shape = NA) +
  aes(x=factor(sampletype, level = c("Midgut", "Ovary", "Larval Water", "Larvae", "Adult")), y=shannon_entropy, fill=sampletype) +
  scale_fill_manual(values = c("#7CC0A6", "#B796C3", "#999EBF", "#8E3F57",  "#C47588"))+
  #coord_cartesian(ylim=c(0,1)) + # adjust y-axis
  xlab("Sample Types") +
  ylab("Shannon's Index")  +
  theme_pubr() +
  geom_jitter(width = 0.2, size=2.5) +
  theme(legend.position="none")+
  facet_grid(~`location`)#remove the legend as it isn't needed

#ggsave("LB_shannon_sampletype_location1007.pdf", height=5, width=9, device="pdf") # save a PDF 5 inches by 9 inches

```



# statistcal analysis of shannon diversity

Are there statistical differences between lab and field data?

```{r}
#Use kruskal wallis, a non-parametric test for global statistic 
kruskal.test(shannon_entropy ~ location, data = shannon_meta) # p-value < 2.2e-16

```

No difference between lab vs field. 

So calculate between sample types within each location


```{r}
# field

shannon_meta_field <- subset(shannon_meta, shannon_meta$location == "Field")

#Use kruskal wallis, a non-parametric test for global statistic 
kruskal.test(shannon_entropy ~ sampletype, data = shannon_meta_field) # p-value < 2.2e-16

```

there are, so Dunn's test to see where differences are

```{r}

dunn.test(shannon_meta_field$shannon_entropy, shannon_meta_field$sampletype,method = "bonferroni") 

```

With low sample sizes the only statistically significant difference in adult vs larvae. Try with just the lab derived lines.


```{r}
# lab

shannon_meta_lab <- subset(shannon_meta, shannon_meta$location == "Lab")

#Use kruskal wallis, a non-parametric test for global statistic 
kruskal.test(shannon_entropy ~ sampletype, data = shannon_meta_lab) # p-value < 2.2e-16

```

there are, so Dunn's test to see where differences are

```{r}

dunn.test(shannon_meta_lab$shannon_entropy, shannon_meta_lab$sampletype,method = "bonferroni") 

```

With the larger sample size here, there are now significant difs between:

larvae vs adults
larval water vs adults
ovary vs larvae
ovary vs larval water


```{r}
#Use kruskal wallis, a non-parametric test for global statistic 
kruskal.test(shannon_entropy ~ sampletype, data = shannon_meta) # p-value < 2.2e-16

```
where are the differences?

```{r}

dunn.test(shannon_meta$shannon_entropy, shannon_meta$sampletype,method = "bonferroni") 

```
this has the extra difference of midgut to larvae.





# beta diversity

## metric 1 - bray curtis 

```{r}

bray <- read_qza("bray_curtis_pcoa_results.qza")
bray <- bray$data$Vectors %>%
  select(SampleID, PC1, PC2) 

bray_meta <- merge(metadata, bray, by="SampleID") # this way I dont need to filter after as it only keeps rows in both original dfs

```


```{r}
ggplot(bray_meta, aes(x=PC1, y=PC2))+geom_point(aes(fill=sampletype), 
       colour="black",pch=21, size=7)+
  ggtitle("Bray-Curtis")+
  scale_fill_manual(values = c("#7CC0A6", "#B796C3", "#999EBF", "#8E3F57",  "#C47588"))+
  theme_bw()+
  xlab("PC1 (23.01%)")+
  ylab("PC2 (13.67%)")+
  labs_pubr()+
  theme(panel.grid.major = element_blank(),
    #    d.minor = element_blank(),
        legend.title = element_blank(),
        legend.text = element_text(size = 16),
        axis.text = element_text(size = 16)) +
    facet_grid(~`location`)

# PC1 and PC2 percentages are from emperor plot

## ggsave("LB_bray-field_lab.pdf", height=5, width=9, device="pdf") save a PDF 5 inches by 9 inches
```


## statistical analysis of bray curtis dissimilarity

```{r}
brayDist <- read_qza("bray_curtis_distance_matrix.qza")
brayDist <- brayDist$data
head(brayDist)

```

```{r}
#make a header for the first column/rownames of the dist matrix 

#To ensure that I'm correctly pairing the sample_type with the order of the brayDist object,
#Make a df that joins the dist and the unordered metadata object 
brayDist_df <- as.matrix(brayDist)
brayDist_df <- as.data.frame(brayDist_df)
brayDist_df <- cbind(rownames(brayDist_df), data.frame(brayDist_df, row.names =NULL))
colnames(brayDist_df)[1] <- "SampleID"
colnames(metadata)[1]<-"SampleID"
joined <- inner_join(brayDist_df,metadata, by = "SampleID")

#Test if it's different by sample_type 
adonis2(brayDist~joined$sampletype)
```

Yep! sampletype does affect beta diversity, now, which comparisons exactly

```{r}
pairwise.adonis2(brayDist~sampletype, data = joined)
```

all pairs show statistically significant differences apart from midgut vs ovary

Are the two locations different? Don't suppose you need this here as there are only 2 locations

```{r}
#Test if it's differnt by location 
adonis2(brayDist~joined$location)

```
```{r}
pairwise.adonis2(brayDist~location, data = joined)
```

# beta diversity - unweighted unifrac

get data

```{r}
uwunifrac <- read_qza("unweighted_unifrac_pcoa_results.qza")
uwunifrac <- uwunifrac$data$Vectors %>%
  select(SampleID, PC1, PC2) 

uwunifrac_meta <- merge(metadata, uwunifrac, by="SampleID") # this way I dont need to filter after as it only keeps rows in both original dfs

```

```{r}
ggplot(uwunifrac_meta, aes(x=PC1, y=PC2))+geom_point(aes(fill=sampletype), 
       colour="black",pch=21, size=7)+
  ggtitle("Unweighted Unifrac")+
  scale_fill_manual(values = c("#7CC0A6", "#B796C3", "#999EBF", "#8E3F57",  "#C47588"))+
  theme_bw()+
  xlab("PC1 (14.36%)")+
  ylab("PC2 (10.37%)")+
  labs_pubr()+
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.title = element_blank(),
        legend.text = element_text(size = 16),
        axis.text = element_text(size = 16))+
    facet_grid(~`location`)

#ggsave("LB_uwunifrac_field_lab.pdf", height=5, width=9, device="pdf") 


# PC1 and PC2 labels came from emperor plot
```


## statistical analysis of unweighted unifrac dissimilarity

```{r}
uwuDist <- read_qza("unweighted_unifrac_distance_matrix.qza")
uwuDist <- uwuDist$data
head(uwuDist)

```

```{r}
#make a header for the first column/rownames of the dist matrix 

#To ensure that I'm correctly pairing the sample_type with the order of the brayDist object,
#Make a df that joins the dist and the unordered metadata object 
uwuDist_df <- as.matrix(uwuDist)
uwuDist_df <- as.data.frame(uwuDist_df)
uwuDist_df <- cbind(rownames(uwuDist_df), data.frame(uwuDist_df, row.names =NULL))
colnames(uwuDist_df)[1] <- "SampleID"
colnames(metadata)[1]<-"SampleID"
joined2 <- inner_join(uwuDist_df,metadata, by = "SampleID")


#Test if it's different by sample_type 
adonis2(uwuDist~joined2$sampletype)# F = 10.929, p = 0.001, R2 = 0.1603

```


```{r}
pairwise.adonis2(uwuDist~sampletype, data = joined2) 

```
with this metric there is more overlap between sampletypes. There is no statistically significant difference between ovary and midgut again, but also between Adult and Ovary or midgut.

Is there a statistically significance between lab and field derived isofemale lines?

```{r}
#Test if it's differnt by location 
adonis2(uwuDist~joined2$location)# F = 3.3046, p = 0.001, R2 = 0.16393

```

```{r}
pairwise.adonis2(uwuDist~location, data = joined2) 

```

# now to separate lab and field samples and look at beta diversity separately

With the alpha diversity I could just separate the data in R and make 2x plots and 2x analyses of statistical differences, but for beta diversity I need to recalculate the differences between samples after subsetting, so doing this in qiime2 with the following code:

$ qiime feature-table filter-samples --i-table ./core-metrics-results-1007/rarefied_table.qza --m-metadata-file metadata_updated_01092022_NAs_removed.txt --p-where "[location]='Field'" --o-filtered-table rarefied-table-Field.qza

$ qiime diversity core-metrics-phylogenetic --i-phylogeny rooted-tree.qza --i-table rarefied-table-Field.qza --p-sampling-depth 1007 --m-metadata-file metadata_updated_01092022_NAs_removed.txt --output-dir core-metrics-results-1007-Field

I subset the already rarefied sample so that these data directly correspond to the abundace/heatmap data.


## metric 1 - bray curtis - lab

```{r}

braylab <- read_qza("bray_curtis_pcoa_results_Lab.qza")
braylab <- braylab$data$Vectors %>%
  select(SampleID, PC1, PC2) 

braylab_meta <- merge(metadata, braylab, by="SampleID") 
```


```{r}
ggplot(braylab_meta, aes(x=PC1, y=PC2))+geom_point(aes(fill=sampletype), 
       colour="black",pch=21, size=7)+
  ggtitle("Bray-Curtis - Lab")+
  scale_fill_manual(values = c("#7CC0A6", "#B796C3", "#999EBF", "#8E3F57",  "#C47588"))+
  theme_bw()+
  xlab("PC1 (24.91%)")+
  ylab("PC2 (14.01%)")+
  labs_pubr()+
  theme(panel.grid.major = element_blank(),
    #    d.minor = element_blank(),
        legend.title = element_blank(),
        legend.text = element_text(size = 16),
        axis.text = element_text(size = 16)) 

# PC1 and PC2 percentages are from emperor plot

## ggsave("LB_bray_lab.pdf", height=5, width=9, device="pdf") save a PDF 5 inches by 9 inches
```

```{r}
braylabDist <- read_qza("bray_curtis_distance_matrix_Lab.qza")
braylabDist <- braylabDist$data
head(braylabDist)

#To ensure that I'm correctly pairing the sample_type with the order of the brayDist object,
#Make a df that joins the dist and the unordered metadata object 
braylabDist_df <- as.matrix(braylabDist)
braylabDist_df <- as.data.frame(braylabDist_df)
braylabDist_df <- cbind(rownames(braylabDist_df), data.frame(braylabDist_df, row.names =NULL))
colnames(braylabDist_df)[1] <- "SampleID"
colnames(metadata)[1]<-"SampleID"
joinedlab <- inner_join(braylabDist_df,metadata, by = "SampleID")

#Test if it's different by sample_type 
adonis2(braylabDist~joinedlab$sampletype)
```

Yep! sampletype does affect beta diversity, now, which comparisons exactly

```{r}
pairwise.adonis2(braylabDist~sampletype, data = joinedlab)
```

Result: ovary and midgut do stat signif difference, all others dif from each other in pairwise comps (all ***, apart from larval water vs midgut thats 0.003)

Now for testing whether dispersal is the same or different amongst sampletypes

```{r}
joinedlab$sampletype <- as.vector(joinedlab$sampletype)
ps.disper_braylab <- betadisper(braylabDist,joinedlab$sampletype) 
ps.disper_braylab 
```

posthoc test

```{r}
permutest(ps.disper_braylab, pairwise = TRUE)  
```

```{r}
summary(permustats(permutest(ps.disper_braylab, pairwise=TRUE)))
```
There are differences, but not all are different from each other, will think about how to word and interpret.



## metric 1 - bray curtis - field
repeating as above but with the field derived isofemale lines data

```{r}

brayfield <- read_qza("bray_curtis_pcoa_results_Field.qza")
brayfield <- brayfield$data$Vectors %>%
  select(SampleID, PC1, PC2) 

brayfield_meta <- merge(metadata, brayfield, by="SampleID") 

ggplot(brayfield_meta, aes(x=PC1, y=PC2))+geom_point(aes(fill=sampletype), 
       colour="black",pch=21, size=7)+
  ggtitle("Bray-Curtis - Field")+
  scale_fill_manual(values = c("#7CC0A6", "#B796C3", "#999EBF", "#8E3F57",  "#C47588"))+
  theme_bw()+
  xlab("PC1 (32.67%)")+
  ylab("PC2 (13.46%)")+
  labs_pubr()+
  theme(panel.grid.major = element_blank(),
    #    d.minor = element_blank(),
        legend.title = element_blank(),
        legend.text = element_text(size = 16),
        axis.text = element_text(size = 16)) 

# PC1 and PC2 percentages are from emperor plot

## ggsave("LB_bray_field.pdf", height=5, width=9, device="pdf") save a PDF 5 inches by 9 inches
```


# statistical analyses

```{r}
brayfieldDist <- read_qza("bray_curtis_distance_matrix_Field.qza")
brayfieldDist <- brayfieldDist$data

brayfieldDist_df <- as.matrix(brayfieldDist)
brayfieldDist_df <- as.data.frame(brayfieldDist_df)
brayfieldDist_df <- cbind(rownames(brayfieldDist_df), data.frame(brayfieldDist_df, row.names =NULL))
colnames(brayfieldDist_df)[1] <- "SampleID"
colnames(metadata)[1]<-"SampleID"
joinedfield <- inner_join(brayfieldDist_df,metadata, by = "SampleID")


adonis2(brayfieldDist~joinedfield$sampletype)
```
```{r}
pairwise.adonis2(brayfieldDist~sampletype, data = joinedfield)
```
less statistical significance than the lab samples where there are more samples, but the same pattern.

Now for testing dispersal 

```{r}
joinedfield$sampletype <- as.vector(joinedfield$sampletype)
ps.disper_brayfield <- betadisper(brayfieldDist,joinedfield$sampletype) 
ps.disper_brayfield 
```

posthoc test

```{r}
permutest(ps.disper_brayfield, pairwise = TRUE) 
summary(permustats(permutest(ps.disper_brayfield, pairwise=TRUE)))
```



## metric 2 - unweighted unifrac -lab

```{r}

uw_uni_lab <- read_qza("unweighted_unifrac_pcoa_results_Lab.qza")
uw_uni_lab <- uw_uni_lab$data$Vectors %>%
  select(SampleID, PC1, PC2) 

uw_uni_lab_meta <- merge(metadata, uw_uni_lab, by="SampleID") 

ggplot(uw_uni_lab_meta, aes(x=PC1, y=PC2))+geom_point(aes(fill=sampletype), 
       colour="black",pch=21, size=7)+
  ggtitle("Unweighted unifrac - Lab")+
  scale_fill_manual(values = c("#7CC0A6", "#B796C3", "#999EBF", "#8E3F57",  "#C47588"))+
  theme_bw()+
  xlab("PC1 (21.03%)")+
  ylab("PC2 (6.79%)")+
  labs_pubr()+
  theme(panel.grid.major = element_blank(),
    #    d.minor = element_blank(),
        legend.title = element_blank(),
        legend.text = element_text(size = 16),
        axis.text = element_text(size = 16)) 

# PC1 and PC2 percentages are from emperor plot

## ggsave("LB_uw-uni_lab.pdf", height=5, width=9, device="pdf") save a PDF 5 inches by 9 inches
```

```{r}
uwlabDist <- read_qza("unweighted_unifrac_distance_matrix_Lab.qza")
uwlabDist <- uwlabDist$data

uwlabDist_df <- as.matrix(uwlabDist)
uwlabDist_df <- as.data.frame(uwlabDist_df)
uwlabDist_df <- cbind(rownames(uwlabDist_df), data.frame(uwlabDist_df, row.names =NULL))
colnames(uwlabDist_df)[1] <- "SampleID"
colnames(metadata)[1]<-"SampleID"
uwjoinedlab <- inner_join(uwlabDist_df,metadata, by = "SampleID")

adonis2(uwlabDist~uwjoinedlab$sampletype)
```
```{r}
pairwise.adonis2(uwlabDist~sampletype, data = uwjoinedlab)
```
and dispersal
```{r}
uwjoinedlab$sampletype <- as.vector(uwjoinedlab$sampletype)
ps.disper_uwlab <- betadisper(uwlabDist,uwjoinedlab$sampletype) 
ps.disper_uwlab 
```

```{r}
permutest(ps.disper_uwlab, pairwise = TRUE) 
summary(permustats(permutest(ps.disper_uwlab, pairwise=TRUE)))
```


## metric 2 - unweighted unifrac -field




```{r}

uw_uni_field <- read_qza("unweighted_unifrac_pcoa_results_Field.qza")
uw_uni_field <- uw_uni_field$data$Vectors %>%
  select(SampleID, PC1, PC2) 

uw_uni_field_meta <- merge(metadata, uw_uni_field, by="SampleID") 

ggplot(uw_uni_field_meta, aes(x=PC1, y=PC2))+geom_point(aes(fill=sampletype), 
       colour="black",pch=21, size=7)+
  ggtitle("Unweighted unifrac - Field")+
  scale_fill_manual(values = c("#7CC0A6", "#B796C3", "#999EBF", "#8E3F57",  "#C47588"))+
  theme_bw()+
  xlab("PC1 (17.48%)")+
  ylab("PC2 (11.38%)")+
  labs_pubr()+
  theme(panel.grid.major = element_blank(),
    #    d.minor = element_blank(),
        legend.title = element_blank(),
        legend.text = element_text(size = 16),
        axis.text = element_text(size = 16)) 

# PC1 and PC2 percentages are from emperor plot

## ggsave("LB_uw-uni_field.pdf", height=5, width=9, device="pdf") save a PDF 5 inches by 9 inches
```

```{r}
uwfieldDist <- read_qza("unweighted_unifrac_distance_matrix_Field.qza")
uwfieldDist <- uwfieldDist$data

uwfieldDist_df <- as.matrix(uwfieldDist)
uwfieldDist_df <- as.data.frame(uwfieldDist_df)
uwfieldDist_df <- cbind(rownames(uwfieldDist_df), data.frame(uwfieldDist_df, row.names =NULL))
colnames(uwfieldDist_df)[1] <- "SampleID"
colnames(metadata)[1]<-"SampleID"
uwjoinedfield <- inner_join(uwfieldDist_df,metadata, by = "SampleID")

adonis2(uwfieldDist~uwjoinedfield$sampletype)
```

```{r}
pairwise.adonis2(uwfieldDist~sampletype, data = uwjoinedfield)
```

and dispersal
```{r}
uwjoinedfield$sampletype <- as.vector(uwjoinedfield$sampletype)
ps.disper_uwfield <- betadisper(uwfieldDist,uwjoinedfield$sampletype) 
ps.disper_uwfield 
```


```{r}
permutest(ps.disper_uwfield, pairwise = TRUE) 
summary(permustats(permutest(ps.disper_uwfield, pairwise=TRUE)))
```

## tests with seed setting

```{r}
set.seed(1234)
permutest(ps.disper_uwfield, pairwise = TRUE) 
summary(permustats(permutest(ps.disper_uwfield, pairwise=TRUE)))
```


```{r}
set.seed(1234)
permutest(ps.disper_uwfield, pairwise = TRUE) 
summary(permustats(permutest(ps.disper_uwfield, pairwise=TRUE)))
```




