---
title: "LB_5_make_save_phyloseq_obj"
author: "Laura Brettell"
date: "2022-11-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The project

Vertical transmission – isofemale lines and rearing in closed conditions free from external microbes to examine transfer of bacteria between generations via the egg surface.

## About

This script makes a phyloseq object after filtering in r and qiime2



# Install packages

```{r}
# phyloseq itself
if(!requireNamespace("BiocManager")){
  install.packages("BiocManager")
   }
BiocManager::install("phyloseq")

# qiime2r
# to be able to convert my qiime2 files
#if (!requireNamespace("devtools", quietly = TRUE)){install.packages("devtools")}
# devtools::install_github("jbisanz/qiime2R") # current version is 0.99.20
 
# decontam for filtering contaminant reads
# if (!require("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")

#BiocManager::install("decontam")

#install.packages("MetBrewer")


#library(MetBrewer)
library(phyloseq)
library(qiime2R)
#library(ggplot2)
#library(tidyverse)
library(dplyr) # for select command
#library(decontam)
#library(vegan)
#library(gridExtra) # to arrange heatmaps


```



# ASV

```{r}
# get taxa
ASVs <- read_qza("./rarefied_table.qza") # now read in ASV data
ASVs <- as.data.frame(ASVs$data)
OTU <- otu_table(ASVs, taxa_are_rows = TRUE)

list_of_ASVs <- row.names(ASVs) # for subsetting tax table

list_of_samples <- colnames(ASVs) # for subsetting metadata
```

# taxonomy

```{r}
# get taxonomy info
taxonomydata <- read_qza("./taxonomy-silva.qza")$data # read in data
tax_parse <- parse_taxonomy(taxonomydata)# make in more useful format

# this is taxonomy for all ASVs, I only want those that are left after rarefying
taxonomy_4516 <- tax_parse[list_of_ASVs, ]
taxonomy_4516 <- as.matrix(taxonomy_4516)

TAX <- tax_table(taxonomy_4516)

```


# metadata


```{r}
metadata <- read_q2metadata("./metadata_updated_01092022_NAs_removed_new_cols.txt") # this is everything including those samples which were removed after rarefaction etc

metadata_subset <- filter(metadata, SampleID %in% c(list_of_samples))

```


# put into phyloseq object 



```{r}

# create phyloseq object
physeq = phyloseq(OTU, TAX) #create phyloseq object
physeq # look at it
sample_names(physeq) # check

# get metadata matching up correctly
physeq_name_order <- sample_names(physeq) # need to get the samples in the same order as they appear in the physeq object (ie the column orders of the ASVs)

metadata_physeq_order <- metadata_subset 

metadata_physeq_order %>%  arrange(SampleID, physeq_name_order)

# make object of a class that phyloseq understands, by cropping the columns of interest from the metadata and specifying the same rows as the original phloseq object

sampledata = sample_data(data.frame(
  Sample = metadata_physeq_order$SampleID,
  seqID = metadata_physeq_order$seqid,
  location = metadata_physeq_order$location,
  line = metadata_physeq_order$line,
  sampletype = metadata_physeq_order$sampletype,
  neg_control = metadata_physeq_order$neg_control,
  row.names=sample_names(physeq),
  stringsAsFactors=FALSE
))

physeq1007 = merge_phyloseq(physeq, sampledata)
physeq1007


```

# save phyloseq object so I dont need all the commands every time

```{r}

#Save the final phyloseq object here: 

saveRDS(physeq1007, "physeq1007")
#use this command to save the phyloseq object to the working directory
#So that you won't have to run all the commands above each time 

ps = readRDS("physeq1007")
#run this to load the  phyloseq obj 
ps

```


